A Bayesian ranking statistic to find \\high-mass black holes in LIGO data

### 2021-01-28 main.tex

\newcommand{\pastro}{\fancytext{p_\text{astro}}}
\newcommand{\pastrobcr}{\fancytext{p_\text{astro}^{\text{BCR}}}}
\newcommand{\pastroext}{\fancytext{p_\text{astro}^{\text{ext}}}}
\newcommand{\pval}{\fancytext{\text{p-value}}}
% Table settings
\renewcommand{\aboverulesep}{0pt}
\renewcommand{\belowrulesep}{0pt}
% \setlength\cellspacetoplimit{2pt}
% \setlength\cellspacebottomlimit{2pt}

@@

\input{Data/results_table}
We analyse the $60\ 996$ background, $5\ 146$ simulated, and $25$ foreground triggers reported by \pycbc's search of the data from the second observing run, restricting our analysis to the triggers that fall within our Bayesian prior space as described in Section~\ref{sec:method}. In addition to these triggers, we also analyse events and candidate events reported by GWTC-1 and the IAS group (note that some of these were identified as candidates by the \pycbc search). In Table~\ref{tab:results}, we summarise the \pastrobcr, along with the \pastro of other pipelines for comparison. Note that although the various pipeline \pastro are not mathematically equivalent, by comparing pipeline \pastro values for a given event, we can compare how significant each pipeline deems various candidates.

@@

\subsection{IAS Events}
Our analysis of the \av{double check that these fall in our search space!!} high-mass IAS events and candidates in O2 has resulted in three events with disfavourable $\pastrobcr<0.5$ (GW170202, GW170403, GW170425). On the other hand, five events (GW170121, GW170304, GW170402, GW170727, GW170817A) and one candidate event (170302) reported by IAS have $\pastrobcr>0.5$.

GW170727 and 170302 are of particular interest, with \pastrobcr of 0.94 and 0.73. GW170727 was emitted from a black hole binary system with a source frame total mass $\approx 70\ \msun$. In addition to the high \pastrobcr reported by our study, \citet{IAS1} and \citet{pycbc_ogc_2} have also reported high \pastro values of 0.98 and 0.99, making it a viable gravitational-wave event candidate. Similarly, the sub-marginal-candidate 170302 reported by \cite{IAS1} with a \pastro of 0.45 appears to have a higher significance from our analysis, resulting in a \pastrobcr of $0.73$.

\av{GW170202's component black holes have source masses (23, 16) below that of our prior space which may be why it has a low \pastrobcr}

\av{Interesting that GW170817A has high \pastrobcr as IAS2's pipeline works well on single detector events (marginal in other detectors)}

\av{why do IAS2:GW170425 and IAS2:GW170402 have low \pastro?}

\subsection{New Candidate Events}
Although no clear detections are made with the \bcr, a marginal-candidate 170222 has been discovered with a \pastrobcr of $0.53$. This candidate has its similar masses when compared to those of GWTC-1. The other candidates with \pastrobcr

@@

Although our analysis targets high-mass triggers, this method can be extended to include the entire body range of LIGO-detectable gravitational wave sources. Additionally, to further improve the method's infrastructure, we can use more robust gravitational wave templates (such as templates that incorporate higher-order modes), and sophisticated glitch models. Future analysis can also incorporate data from all available detectors in a network to increase the \pastro's sensitivity. The BCR can discern better whether a candidate is a coherent astrophysical candidate or an incoherent glitch with data from more detectors.

@@

\section{Tuning signal and glitch prior odds}

Table of the signal and glitch prior odds utilised for each section of data. One KL plot

\section{Posteriors}

>>>>

\newcommand{\pastro}{\fancytext{p_\text{astro}}}
\newcommand{\pastrobcr}{\fancytext{p_\text{astro}^{\text{BCR}}}}
\newcommand{\untunedpastrobcr}{\fancytext{p_\text{astro}^{\text{BCR}\prime}}}
\newcommand{\pastroext}{\fancytext{p_\text{astro}^{\text{ext}}}}
\newcommand{\pval}{\fancytext{\text{p-value}}}
% Table settings
\renewcommand{\aboverulesep}{0pt}
\renewcommand{\belowrulesep}{0pt}
% \setlength\cellspacetoplimit{2pt}
% \setlength\cellspacebottomlimit{2pt}

@@

\input{Data/results_table}
We analyse the $60\ 996$ background, $5\ 146$ simulated, and $25$ candidate triggers reported by \pycbc's search of the data from the second observing run, restricting our analysis to the triggers that fall within our Bayesian prior space as described in Section~\ref{sec:method}. In addition to these triggers, we also analyse events and candidate events reported by GWTC-1 and the IAS group (note that some of these were identified as candidates by the \pycbc search). In Table~\ref{tab:results}, we summarise the \pastrobcr, along with the \pastro of other pipelines for comparison. Note that although the various pipeline \pastro are not mathematically equivalent, by comparing pipeline \pastro values for a given event, we can compare how significant each pipeline deems various candidates. The $\alpha$ and $\beta$ values utilised for each time-frame of O2 are reported in Appendix~\ref{apdx:alphabeta}

@@

\subsection{IAS Events}
Our analysis of the high-mass IAS events and candidates in O2 has resulted in three events with disfavourable $\pastrobcr<0.5$ (GWC170402, GW170403, GW170425), and four events and one candidate with $\pastrobcr\geq0.5$ (GW170121, 170302, GW170304, GW170727, GW170817A).

GWC170402, detected by \citet{IAS2}, is reported to have a signal that is not described well by waveforms for circular binaries with aligned spins. Hence, we might have received a low \pastrobcr due to our usage of \imrphenomp, a waveform that does not account for eccentricity. Additionally, the search conducted by \citet{IAS2} was a single-detector search. Our ranking statistic relies on the signal to appear coherent, even if just faintly coherent, amongst the various detectors to have a high \pastrobcr. Hence, the lack of coherence and the non-eccentric waveform may be the leading factors for a low \pastro. GW170403 and GW170425 which have $\pastrobcr<0.35$ also have low  \pastro reported by \citet{pycbc_ogc_2},  suggesting that these events may have been false alarms.

From the candidates with $\pastrobcr>0.5$, GW170727 and 170302 are of particular interest, with \pastrobcr of $0.92$ and $0.63$. GW170727 was emitted from a black hole binary system with a source frame total mass $\approx 70\ \msun$. In addition to the high \pastrobcr reported by our study, \citet{IAS1} and \citet{pycbc_ogc_2} have also reported high \pastro values of 0.98 and 0.99, making it a viable gravitational-wave event candidate. Similarly, the sub-marginal-candidate 170302 reported by \cite{IAS1} with a \pastro of 0.45 appears to have a higher significance from our analysis, resulting in a \pastrobcr of $0.63$.

% \av{Interesting that GW170817A has high \pastrobcr as IAS2's pipeline works well on single detector events (marginal in other detectors)}

\subsection{New Candidate Events}
Although no clear detections are made with the \bcr, a marginal-candidate 170222 has been discovered with a $\pastrobcr\sim0.5$. This candidate has its similar masses when compared to those of GWTC-1. The remaining coherent trigger candidates all have $\pastrobcr\ll0.5$ making them unlikely to originate from astrophysical sources.

@@

Although our analysis targets high-mass triggers, this method can be extended to include the entire body range of LIGO-detectable gravitational wave sources. Additionally, to further improve the method's infrastructure, we can use more robust gravitational wave templates (such as templates that incorporate higher-order modes), and sophisticated glitch models. Future analysis can also incorporate data from all available detectors in a network to increase the sensitivity of \pastrobcr. The BCR can discern better whether a candidate is a coherent astrophysical candidate or an incoherent glitch with data from more detectors.

@@

\section{Tuned prior odds}\label{apdx:alphabeta}

O2 lasted several months over which the detector's sensitivity varied.  Hence, a part of our analysis entailed tuning the prior odds for obtaining a signal and a glitch, $\alpha$ and $\beta$, as described in Section~\ref{sec:method}. Table~\ref{tab:priorodds} presents the signal and glitch prior odds utilised for each time-frame of O2 data.
\input{Data/alpha_beta_table}

Tuning the prior odds can dramatically affect the \pastrobcr. For example, consider Table~\ref{tab:tuningresults}, which reports tuned \pastrobcr and un-tuned \untunedpastrobcr (where $\alpha=1$ and $\beta=1$) for various high-mass events and candidates. By tuning the prior odds, the \pastrobcr for some IAS events (for example, GW170403 and GW170817A) can change by more than 0.5, resulting in a candidate being promoted/demotion of candidate significance. 
\input{Data/tuning_results_table}
